### 2019-02-19 sections/intuition.tex
clarify difference between shared data and shared editor
removed three qed's
fixes in intuition

The central objective of \TOP is to \emph{coordinate collaboration}.
The basic building blocks of \TOPHAT are editors and combinators.
Editors are tasks with which users interact directly.
Combinators express ways in which people can work together.
Tasks can be executed after each other, at the same time, or conditionally.
This motivates the combinators step, parallel, and choice.

@@

\begin{example}[Twice]

A higher-order function which executes a given task twice.
\begin{TASK}
  let twice : Task $\tau$ -> Task $\tau$ = \t : Task $\tau$. t >>? \x . t
\end{TASK}
The task value of the first execution is ignored.
\qed

@@

Input events drive evaluation of tasks.
When the system receives a valid event, it applies the event to the current task, which results in a new task.
In this way the system communicates with the environment.

In \TOPHAT, editors are the basic method of communication with the environment.
There are different editors, denoted by different box symbols.
An editor of the form $\Edit v$, where $v$ is of type $\tau$, reacts to two kinds of events:
Change events $v$ with a value of type $\tau$ and clear events $\Empty$ which empty the editor.

@@

This example demonstrates external communication and choice.
It is a vending machine that dispenses a small biscuit for one coin and a large biscuit for two coins.
\begin{TASK}
  let vend : Task String = enter Int >>? \n. if n == 1 then edit "small"
    else if n == 2 then edit "large"
    else fail
\end{TASK}
The editor $\Enter \Int$ asks the user to enter an amount of money.
This editor stands for a coin slot in a real machine that freely accepts and returns coins.
There is a continue button that is initially disabled, due to the fact that the left hand side of the step combinator has no value.
When the user has inserted exactly 1 or 2 coins, the continue button becomes enabled.
When the user presses the continue button, the machine dispenses either a large or a small biscuit, depending on the amount of money. \qed

@@

The final observation is to determine whether a task results in $\Fail$.
The step combinator and the choice combinator use this to prevent users from picking a failing task.

@@

In a step $t \Next e$, the decision to move on is taken by $\Next$, not by $t$.
The decision is based on a speculative evaluation of $e$.
The step combinators in $t \Then e$ and $t \Next e$ pass the value $v$ of $t$ to the continuation $e$.
Both steps act like $t$ as long as the step is guarded.
A step is guarded if either the left task has no value, or the speculative evaluation $e\ v$ yields $\Fail$.
Once it becomes unguarded, the step continues as $e\ v$.
The task $t \Next e$ additionally requires a continue event $\Continue$ to proceed.
Speculative evaluation is designed so that it does not have side effects.

@@

The step combinator is one form of internal communication, where task values are passed to continuations.
Another form of internal communication is shared data.
Shared data enables data flow \emph{across} control flow, in particular between parallel tasks.
Shared data sources are assignable references whose changes are immediately visible to all tasks interested in them.
If $r$ is a reference of type $\tau$, then $\Update r$ is an editor whose value is that of $r$.

@@

\TOPHAT allows a simple solution to this problem, using guarded steps.
Steps can be guarded with arbitrary expressions, and the parallel combinator can be used to watch two shared editors at the same time.
Let \TS{match}, \TS{paper}, and \TS{tobacco} be references to Booleans.
The smokers are defined as follows.
\begin{TASK}
  let continue = \<<x,y>>. if x /\ y then smoke else fail in
  let tobaccoSmoker = (update match <&> update paper) >>? continue in
  let paperSmoker = (update tobacco <&> update match) >>? continue in
  let matchSmoker = (update tobacco <&> update paper) >>? continue
\end{TASK}
When the agent supplies two of the ingredients by setting the respective shares to \TS{True}, only the step of the smoker that waits for those becomes enabled. \qed

@@

The parallel combinator $\And$ introduces concurrency: Two things are done at the same time.
Let $t_1$ and $t_2$ be tasks.
The parallel combination $t_1 \And t_2$ stands for two independent tasks carried out at the same time.
For the system it does not matter if the tasks are executed by two people actually in parallel, or by one person who switches between the tasks.
The inputs sent to the component tasks are interleaved into a serial stream, which is sent to the parallel combinator.
We assume that such a serialization is always possible.
The tasks are truly independent of each other, all interleavings are permitted.
The environment must prefix events to $t_1$ and $t_2$ by $\First$ (first) and $\Second$ (second) respectively.
This unambiguously renames the inputs, removing any source of nondeterminism.

>>>>

The central objective of \TOP is to \emph{coordinate collaboration}.
The basic building blocks of \TOPHAT are editors and combinators.
Editors are tasks with which users interact directly.
Combinators express ways in which people can work together.
Tasks can be executed after each other, at the same time, or conditionally.
This motivates the combinators step ($\Then$, $\Next$), parallel ($\And$), and choice ($\Or$, $\Xor$).

\begin{example}[Breakfast]
\label{exm:breakfast}

The following program shows the different collaboration operators in action.
Users have a choice whether they want tea or coffee.
They always get an egg.
The drink and the food are prepared in parallel.
When drink and food are prepared, users can step to eating the result.

\begin{TASK}
  let makeBreakfast : Task Drink -> Task Food -> Task Unit =
    \makeDrink . \makeFood . makeDrink <&> makeFood >>? \<<d, f>>. eat d f
  let main : Task Unit =
    makeBreakfast (makeTea <?> makeCoffee) makeEgg
\end{TASK}

@@

\Cref{exm:breakfast} demonstrates how tasks can be parameterized by other tasks: makeBreakfast is a collaboration pattern that always works the same way, regardless of which food and drink are being prepared.

@@

Input events drive evaluation of tasks.
When the system receives a valid event, it applies the event to the current task, which results in a new task.
In this way the system communicates with the environment.
Input events are synchronous.
We assume systems react instantaneous to each such event.

In \TOPHAT, editors are the basic method of communication with the environment.
There are different editors, denoted by different box symbols.
An editor of the form $\Edit v$, where $v$ is of type $\tau$, reacts to two kinds of events:
Change events $v$ with a value of type $\tau$ and clear events $\Empty$ which empty the editor.
This results in a typed empty editor $\Enter \tau$, which is annotated with the same type.

@@

This example demonstrates external communication and choice.
It is a vending machine that dispenses a biscuit for one coin and a chocolate bar for two coins.
\begin{TASK}
  let vend : Task Snack = enter Int >>? \n. if n == 1 then edit Biscuit
    else if n == 2 then edit ChocolateBar
    else fail
\end{TASK}
The editor $\Enter \Int$ asks the user to enter an amount of money.
This editor stands for a coin slot in a real machine that freely accepts and returns coins.
There is a continue button that is initially disabled, due to the fact that the left hand side of the step combinator has no value.
When the user has inserted exactly 1 or 2 coins, the continue button becomes enabled.
When the user presses the continue button, the machine dispenses either a biscuit or a chocolate bar, depending on the amount of money.
Snacks are modelled using a custom data type.

@@

The final observation is to determine whether a task results in $\Fail$ (failure).
The step combinator and the choice combinator use this to prevent users from picking a failing task.

@@

In a step $t \Next e$, the decision to move on is taken by $\Next$, not by $t$.
The decision is based on a speculative evaluation of $e$.
The step combinators in $t \Then e$ and $t \Next e$ pass the value $v$ of $t$ to the continuation $e$.
Both steps act like $t$ as long as the step is guarded.
A step is guarded if either the left task has no value, or the speculative evaluation $e\ v$ yields $\Fail$.
Once it becomes unguarded, the step continues as $e\ v$.
The task $t \Next e$ additionally requires a continue event $\Continue$ to proceed.
Speculative evaluation is designed so that possible side effects are undone.

@@

The step combinator is one form of internal communication, where task values are passed to continuations.
Another form of internal communication is shared data.
Shared data enables data flow \emph{across} control flow, in particular between parallel tasks.
Shared data sources are assignable references whose changes are immediately visible to all tasks interested in them.
Users can not directly interact with shared data, a shared editor is required for that.
If $r$ is a reference of type $\tau$, then $\Update r$ is an editor whose value is that of $r$.

@@

\TOPHAT allows a simple solution to this problem, using guarded steps.
Steps can be guarded with arbitrary expressions, and the parallel combinator can be used to watch two shared editors at the same time.
Let \TS{match}, \TS{paper}, and \TS{tobacco} be references to Booleans.
The smokers are defined as follows.
\begin{TASK}
  let continue = \<<x,y>>. if x /\ y then smoke else fail in
  let tobaccoSmoker = (update match <&> update paper) >>? continue in
  let paperSmoker = (update tobacco <&> update match) >>? continue in
  let matchSmoker = (update tobacco <&> update paper) >>? continue
\end{TASK}
When the agent supplies two of the ingredients by setting the respective shares to \TS{True}, only the step of the smoker that waits for those becomes enabled.

@@

Let $t_1$ and $t_2$ be tasks.
The parallel combination $t_1 \And t_2$ stands for two independent tasks carried out at the same time.
Therefore $\And$ introduces concurrency.
For the system it does not matter if the tasks are executed by two people actually in parallel, or by one person who switches between the tasks.
The inputs sent to the component tasks are interleaved into a serial stream, which is sent to the parallel combinator.
We assume that such a serialization is always possible.
The tasks are truly independent of each other, all interleavings are permitted.
The environment must prefix events to $t_1$ and $t_2$ by $\First$ (first) and $\Second$ (second) respectively.
This unambiguously renames the inputs, removing any source of nondeterminism.
